### 2022-04-06 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_betacov_mapping@ce714f31455d6ef4bdd57d7ff199cef40f1e81a6 🚀

This template uses \texttt{pandoc} (and a few additional python glue
scripts) to facilitate the production of scientific articles using a
standard markdown file. The objective is to ensure that standard
markdown (with the important exception of the \texttt{pandoc-crossref}
citation markup) will be rendered into an interactive website (which
allows collaborative annotations with the \texttt{hypothes.is}
platform), a ``draft'' style PDF (double-spaced, numbered lines, figures
at the end), and a ``preprint'' style PDF (with slightly more
reader-friendly pagination).

The core bit of configuration is the \texttt{metadata.json} file, which
handles information about authorship, affiliations, the abstract,
keywords, etc. All documents will be deployed to \texttt{gh-pages}
\emph{only} on push events from the \texttt{main} branch. All of the
artifacts will be built when doing pull requests, so you can check that
merging a branch is \emph{not} going to cause the compilation of the
documents to fail; indeed, you can download the artifacts produced
during the run, to check the PDF and html files. The website is only
updated from the \texttt{main} branch.

The workflow is \emph{very} GitHub based, and so the manuscript file
\emph{is} the \texttt{README.md} - this is not going to be a huge issue
as 90\% of the markdown is standard, with the exception of the citations
and mathematics, so this will render (mostly) like a normal README file.

\hypertarget{deploying-the-template}{%
\section{Deploying the template}\label{deploying-the-template}}

The process of deploying this template has been \emph{greatly}
streamlined from previous versions:

\begin{itemize}
\tightlist
\item
  Click on the ``Use this template'' button
\item
  Edit \texttt{README.md} with your own text, commit, and push
\item
  This push will trigger the first build - the builds are only active on
  the \texttt{main} branch (\emph{not} \texttt{master}!), and on pull
  requests
\item
  Go to \texttt{http://you.github.io/repo-name/} to view the html
  version, and get access to the PDFs
\item
  Add your references to the \texttt{references.bib} file
\item
  Edit the \texttt{metadata.json} file to add the title, abstract,
  authors
\end{itemize}

In particular, note that \emph{you do not need} to create a personnal
access token to deploy to \texttt{gh-pages} (from where the website is
served).

\hypertarget{the-metadata-file}{%
\section{The metadata file}\label{the-metadata-file}}

\hypertarget{general-information}{%
\subsection{General information}\label{general-information}}

The title is a field in the \texttt{metadata.json}:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{\{}
    \DataTypeTok{"title"}\FunctionTok{:} \StringTok{"Preprint template"}
\FunctionTok{\}}
\end{Highlighting}
\end{Shaded}

\hypertarget{authorship}{%
\subsection{Authorship}\label{authorship}}

Authors are listed as objects in the \texttt{authors} block. Each author
is specified as follows:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{\{}
      \DataTypeTok{"familyname"}\FunctionTok{:} \StringTok{"Bob"}\FunctionTok{,}
      \DataTypeTok{"givennames"}\FunctionTok{:} \StringTok{"Alice"}\FunctionTok{,}
      \DataTypeTok{"email"}\FunctionTok{:} \StringTok{"alice.bob@u.edu"}\FunctionTok{,}
      \DataTypeTok{"orcid"}\FunctionTok{:} \StringTok{"0000{-}0000{-}0000{-}0001"}\FunctionTok{,}
      \DataTypeTok{"affiliations"}\FunctionTok{:} \OtherTok{[}
        \StringTok{"Affiliation 1"}\OtherTok{,}
        \StringTok{"Affiliation 2"}
      \OtherTok{]}\FunctionTok{,}
      \DataTypeTok{"status"}\FunctionTok{:} \OtherTok{[}\StringTok{"corresponding"}\OtherTok{,} \StringTok{"equal"}\OtherTok{]}
    \FunctionTok{\}}
\end{Highlighting}
\end{Shaded}

The \texttt{email} field is recommended for all authors. The
\texttt{status} field is only useful for the corresponding author, and
to denote equal contributions. These informations are rendered on the
initial page. If an \texttt{orcid} is given, it will be linked on the
HTML and PDF versions.

Note that there is \emph{no need} to number the affiliations - a small
python script will take care of this automatically.

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

This template supports three types of abstracts, indicated in the
metadata file as \texttt{abstract}:

A regular \texttt{abstract} is defined as

\begin{Shaded}
\begin{Highlighting}[]
\ErrorTok{"abstract":} \ErrorTok{"A} \ErrorTok{very} \ErrorTok{long} \ErrorTok{string"}
\end{Highlighting}
\end{Shaded}

An itemized abstract is an array of strings, each representing a bullet
point:

\begin{Shaded}
\begin{Highlighting}[]
\ErrorTok{"abstract":} \OtherTok{[}
    \StringTok{"Point 1"}\OtherTok{,}
    \StringTok{"Point 2"}
\OtherTok{]}
\end{Highlighting}
\end{Shaded}

A structured abstract is an object with key-value pairs :

\begin{Shaded}
\begin{Highlighting}[]
\ErrorTok{"abstract":} \FunctionTok{\{}
    \DataTypeTok{"Location"}\FunctionTok{:} \StringTok{"Worldwide"}\FunctionTok{,}
    \DataTypeTok{"Organisms"}\FunctionTok{:} \StringTok{"Mammals"}
\FunctionTok{\}}
\end{Highlighting}
\end{Shaded}

\hypertarget{citation-style}{%
\subsection{Citation style}\label{citation-style}}

The \texttt{citationstyle} key corresponds to the name, with
\texttt{.csl} ommited, of a CSL stylesheet stored in the
\href{https://github.com/citation-style-language/}{citation style
language} repository. Note that there is no difference between main and
dependent styles, the build engine will take the correct steps to get
the correct style. The default is
\texttt{"citationstyle":\ "ecology-letters"}. There is a longer section
about references management later on.

\hypertarget{references-management}{%
\section{References management}\label{references-management}}

The references are managed by \texttt{pandoc}. Note that we \emph{do
not} use \texttt{pandoc-citeproc}, which was an external module for
older \texttt{pandoc} versions. References \emph{must} be stored in a
\texttt{references.bib} file, and that it would make sense to order it
alphabetically by key.

We use \href{https://www.zotero.org/}{Zotero} for references management,
and for the lab's manuscripts, we work from folders in a shared library
(with a folder for every manuscript).

It is recommedned to use the
\href{https://retorque.re/zotero-better-bibtex/}{Better BibTeX} plugin
for citation key generations, and auto-export of the shared library to
the \texttt{references.bib} file. We use a citation key format meant to
convey information on the author (first author full name), date (complet
year), and title (first three letters of the first two non-stop words).
It must be set in the Better BibTeX preferences as (you might need to
remove the line changes):

\begin{verbatim}
[auth:fold]
[year]
[title:fold:nopunctordash:skipwords:lower:select=1,1:substring=1,3:capitalize]
[title:fold:nopunctordash:skipwords:lower:select=2,2:substring=1,3:capitalize]
\end{verbatim}

It is a good idea to configure Better BibTeX to auto-export on change,
and to remove a lot of fields that are not strictly speaking required
for references. The list of fields we usually ignore is:

\begin{verbatim}
abstract,copyright,annotation,file,pmid,month,shorttitle,keywords
\end{verbatim}

The citations are done using the normal markdown syntax, where
\texttt{@Elton1927AniEco} produces Elton (1927), and
\texttt{{[}@Camerano1880EquViv{]}} produces (Camerano 1880).

\hypertarget{figures-tables-and-other-floats}{%
\section{Figures, Tables, and other
floats}\label{figures-tables-and-other-floats}}

Note that you can wrap the text of legends for both figures and tables.
This avoids the issue of having very long lines.

\hypertarget{mathematics}{%
\subsection{Mathematics}\label{mathematics}}

The following equation

\begin{equation}\protect\hypertarget{eq:eq1}{}{J'(p) = \frac{1}{\text{log}(S)}\times\left(-\sum p \times \text{log}(p)\right)}\label{eq:eq1}\end{equation}

is produced using

\begin{Shaded}
\begin{Highlighting}[]
\SpecialStringTok{$$J\textquotesingle{}(p) = }\SpecialCharTok{\textbackslash{}frac}\SpecialStringTok{\{1\}\{}\SpecialCharTok{\textbackslash{}text}\NormalTok{\{log\}}\SpecialStringTok{(S)\}}\SpecialCharTok{\textbackslash{}times}\SpecialStringTok{ ... $$}\NormalTok{ \{\#eq:eq1\}}
\end{Highlighting}
\end{Shaded}

and can be referenced using \texttt{@eq:eq1}, which will result in
eq.~\ref{eq:eq1}. Note that because we use \texttt{pandoc-crossref}, the
label ``eq.'' will be generated automatically.

\hypertarget{tables}{%
\subsection{Tables}\label{tables}}

Table legends go on the line after the table itself. To generate a
reference to the table, use \texttt{\{\#tbl:id\}} -- then, in the text,
you can use \texttt{\{@tbl:id\}} to refer to the table. For example, the
table below is tbl.~\ref{tbl:id}. You can remove the \emph{table} in
front by using \texttt{!@tbl:id}, or force it to be capitalized with
\texttt{\textbackslash{}*tbl:id}.

\hypertarget{tbl:id}{}
\begin{longtable}[]{@{}rrrrl@{}}
\caption{\label{tbl:id}This is a table, and its identifier is
\texttt{id} -- we can refer to it using \texttt{\{@tbl:id\}}. Note that
even if the table legend is written below the table itself, it will
appear on top in the PDF document.}\tabularnewline
\toprule
Sepal.Length & Sepal.Width & Petal.Length & Petal.Width &
Species\tabularnewline
\midrule
\endfirsthead
\toprule
Sepal.Length & Sepal.Width & Petal.Length & Petal.Width &
Species\tabularnewline
\midrule
\endhead
5.1 & 3.5 & 1.4 & 0.2 & setosa\tabularnewline
5.0 & 3.6 & 1.4 & 0.2 & setosa\tabularnewline
5.4 & 3.9 & 1.7 & 0.4 & setosa\tabularnewline
\bottomrule
\end{longtable}

\hypertarget{figures}{%
\section{Figures}\label{figures}}

Figures can have a legend -- all figures \emph{must} be in the
\texttt{figures/} folder of the project, as it is also used for the
website. We recommend to use good resolution images, rather than PDFs,
or at least to have multiple versions available.

\begin{verbatim}
![This is the legend of the figure...](figures/figure.png){#fig:figure}
\end{verbatim}

\begin{figure}
\hypertarget{fig:figure}{%
\centering
\includegraphics{figures/figure.png}
\caption{This is the legend of the figure, which will be shown in the
margin in preprint mode, and underneath the figure in draft mode. The
legend can contain references, etc. It is advised to use a resolution of
at least 600dpi for the figures.}\label{fig:figure}
}
\end{figure}

We can now use \texttt{@fig:figure} to refer to fig.~\ref{fig:figure}.

\hypertarget{example-text}{%
\section{Example text}\label{example-text}}

Connectance, defined as the ratio of realized interactions on the total
number of potential interactions, is one of the most common descriptor
of network structure. In a bipartite network with \(T\) species at the
top, and \(B\) at the bottom, having a total of \(L\) interactions, it
is defined as \(Co = L/(T\times B)\). Connectance has a lower bound, as
the network cannot have fewer interactions that the number of species in
its more speciose level -- the minimal connectance is therefore
\(c_m = \text{max}(T,B)\). This makes the connectance of networks of
different sizes difficult to compare, especially since bipartite
networks tends to have a low connectance. For this reason, we used a
corrected version of connectance, defined as

\begin{equation}\protect\hypertarget{eq:cstar}{}{Co^\star=\frac{L-c_m}{T\times B-c_m} \,.}\label{eq:cstar}\end{equation}

\hypertarget{this-is-a-subsection}{%
\subsection{This is a subsection}\label{this-is-a-subsection}}

This takes values between 0 (the network has the minimal number of
interactions) and 1 (all species are connected), but is robust to
variations in species richness.

\hypertarget{this-is-another-subsection}{%
\subsection{This is another
subsection}\label{this-is-another-subsection}}

This takes values between 0 (the network has the minimal number of
interactions) and 1 (all species are connected), but is robust to
variations in species richness.

\hypertarget{some-non-standard-maths}{%
\subsection{Some non-standard maths}\label{some-non-standard-maths}}

The phylogenetic reconstruction of \(\hat{\mathscr{L}}\) and
\(\hat{\mathscr{R}}\) has an associated uncertainty, represented by the
breadth of the uniform distribution associated to each of their entries.
Therefore, we can use this information to assemble a
\emph{probabilistic} metaweb in the sense of
(\textbf{Poisot2016StrPro?}), \emph{i.e.} in which every interaction is
represented as a single, independent, Bernoulli event of probability
\(p\).

Specifically, we have adopted the following approach. For every entry in
\(\hat{\mathscr{L}}\) and \(\hat{\mathscr{R}}\), we draw a value from
its distribution. This results in one instance of the possible left
(\(\hat{\mathscr{l}}\)) and right (\(\hat{\mathscr{r}}\)) subspaces for
the Canadian metaweb. These can be multiplied, to produce one matrix of
real values. Because the entries in \(\hat{\mathscr{l}}\) and
\(\hat{\mathscr{r}}\) are in the same space where \(\mathscr{L}\) and
\(\mathscr{R}\) were originally predicted, it follows that the threshold
\(\rho\) estimated for the European metaweb also applies. We use this
information to produce one random Canadian metaweb,
\(N = \hat{\mathscr{L}}\)\(\hat{\mathscr{R}}' \ge \rho\).

Because the intervals around some trait values can be broad (in fact,
probably broader than what they would actually be, see \emph{e.g.}
\textbf{Garland1999IntPhy?}), we repeat the above process
\(2\times 10^5\) times, which results in a probabilistic metaweb \(P\),
where the probability of an interaction (here conveying our degree of
trust that it exists given the inferred trait distributions) is given by
the number of times where it appears across all random draws \(N\),
divided by the number of samples. An interaction with \(P_{i,j} = 1\)
means that these two species were predicted to interact in all
\(2\times 10^5\) random draws, etc..

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-Camerano1880EquViv}{}%
Camerano, L. (1880). Dell'equilibrio dei viventi merce la reciproca
distruzione. \emph{Atti Della R. Accad. Delle Sci. Torino}, 15,
393--414.

\leavevmode\hypertarget{ref-Elton1927AniEco}{}%
Elton, C.S. (1927). \emph{Animal ecology}. University of Chicago Press.

\end{CSLReferences}

>>>>

Spillover risk is not unidimensional. From the standpoint of an animal
community, i.e.~a pool of possible hosts, there are a multiplicity of
ecological factors that come into play (Plowright et al.~2017). The
global richness of hosts is one such component commonly
mentioned/analysed (Anthony et al.~2017), but there is an argument to be
made that species who are not competent hosts of a specific virus genus
may not factor into this (Plowright et al.~2015), or that species who
are assumed to share viruses at different rates should be weighted
accordingly (Albery et al.~2020). In mammals, key functional traits (for
which phylogeny is a reasonable proxy) are determinants of the spillover
potential (Olival et al.~2017); these include, notably, TK. Finally,
especially when the pool of potential hosts spans the entire globe,
there may be local host pools that are highly unique; not having been
observed in other locations, these can act on the overall risk either by
providing novel contact opportunities, reflecting unique
host-environment combinations (Engering et al.~2013), or facilitating
rapid changes in specialism (Agosta et al.~2010). In the specific case
of generalist pathogens (as betacoronavirus clearly are), there is
conceptual and empirical support to the idea that these community- level
mechanisms are even more important in driving the overall risk (Power
and Mitchell 2004).

Bats are important reservoir hosts for different classes of
microorganisms (Chu 2008, Donaldson 2010, Li 2010), some of which can
threaten human health. Especially concerning is the fact that bats are
reservoirs for a variety of emerging viruses (Calisher 2006), making
balancing the needs for bat conservation and disease prevention a
potentially difficult act, especially in more densely populated areas
(REF). Chiropterans emerged around 64 million years ago and are one of
the most diverse mammalian orders, with an estimated richness of more
than 12000 species, (Peixoto F et al, 2018) and 14325 known species
Simmons \& Cirranello. They exhibit a broad variety of habitat use,
behaviour, and feeding strategies, resulting in their playing an
essential role in the delivery of several ecosystem services (Kasso
2013), including economic benefits. Over two-thirds of bats are either
obligate or facultative insectivorous mammals, therefore playing an
important role in the regulation of insect pests that can affect crops
(Williams-Guillen 2011), and vectors of diseases that put a risk on
human health (Gonsalves 2013 ). Because bats are globally distributed
and have a long evolutionary history, phylogeographic and biogeographic
approaches are required to shed light on the extant distribution of
coevolutionary processes between bats and the pathogens they carry. As a
consequence, not all areas are facing a risk of human spillover, and
those who do might not be facing risks of the same nature and magnitude.

Yet a comprehensive assessment of the risk of spillover of
betacoronaviruses from bat hosts to humans is limited by the fact that
we do not know the full diversity of viruses associated with every bat
species. Predictive models can help fill in some of these gaps, by
recommending hosts based on known host-virus associations BECKER REF. In
this paper, we examine the biogeographic structure of
bats-betacoronaviruses associations, based on a curated dataset of known
and predicted hosts. We turn these associations into a spatially
explicit additive mapping of zoonotic risk components, which reveals
extreme heterogeneity of risk at the global scale; furthermore, we
identify the Amazon and South-Eastern Asia as hotspots of phylogenetic
distinctiveness of betacoronaviruses; surprisingly, current data suggest
that viral sharing between hosts in high in the Amazon and low in
South-Eastern Asia, which has the potential to result in different
evolutionary dynamics between these two regions.

TK summary of the results

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{known-betacoronavirus-hosts}{%
\subsection{Known betacoronavirus
hosts}\label{known-betacoronavirus-hosts}}

We downloaded the βCoV reservoir database from
https://www.viralemergence.org/betacov on Aug.~2021. This database was
assembled by a combination of data mining, literature surveys, and
application of an ensemble recommender system classifying hosts as
either ``Suspected'' or ``Unlikely'' (REF BECKER). The hosts considered
for this study were all hosts with a known record of a betacoronavirus,
and all those with a ``Suspected'' status in the ensemble model. This
resulted in a list of TK TP unique host species.

\hypertarget{bats-occurrences}{%
\subsection{Bats occurrences}\label{bats-occurrences}}

We downloaded the rangemap of every extant bat species that was either
classified as an empirically documented or a suspected host of
beta-coronaviruses (Becker et al.~2020), according to recent IUCN data
(IUCN 2021). The range maps were subsequently rasterized at a resolution
of approximately TK TP. For every pixel in the resulting raster where at
least one bat host of betacoronavirus was present, we extract the
species pool, which was used to calculate the following risk assessment
components: phylogenetic diversity, bat compositional uniqueness, and
predicted viral sharing risk.

\hypertarget{bats-phylogeography}{%
\subsection{Bats phylogeography}\label{bats-phylogeography}}

For every pixel, we measured Faith's Phylogenetic Diversity (Faith 1992)
based on a recent synthetic tree with robust time calibration, covering
about 6000 mammalian species (Upham et al.~2019). Faith's PD measures
the sum of unique branches from an arbitrary root to a set of tips, and
comparatively larger values indicate a more phylogenetic diverse species
pool. We measured phylogenetic diversity starting from the root of the
entire tree (and not from Chiroptera); this bears no consequences on the
resulting values, since all branches leading up to Chiroptera are only
counted one per species pool, and (as we explain when describing the
assembly of the composite risk map), all individual risk components are
ranged in {[}0,1{]}. This measure incorporates a richness component,
which we chose not to correct for; the interpretation of the
phylogenetic diversity is therefore a weighted species richness, that
accounts for phylogenetic over/under-dispersal in some places.

\hypertarget{bats-compositional-uniqueness}{%
\subsection{Bats compositional
uniqueness}\label{bats-compositional-uniqueness}}

For every species pool, we measured its Local Contribution to
Beta-Diversity (Legendre and De Cáceres 2013); LCBD works from a
species-data matrix (traditionally noted as Y), where species are rows
and sites are columns, and a value of 1 indicates occurrence. We
extracted the Y matrix assuming that every pixel represents a unique
location, and following best practices (Legendre and Condit 2019)
transformed it using Hellinger's distance to account for unequal bat
richness at different pixels. The correction of raw community data is
particularly important for two reasons: first, it prevents the artifact
of richer sites having higher importance; second, it removes the effect
of overall species richness, which is already incorporated in the
phylogenetic diversity component. High values of LCBD indicate that the
pixel has a community that is on average more dissimilar in species
composition than what is expected knowing the entire matrix, i.e.~a more
unique community.

\hypertarget{viral-sharing-between-hosts}{%
\subsection{Viral sharing between
hosts}\label{viral-sharing-between-hosts}}

For all bat hosts of betacoronaviruses, we extracted their predicted
viral sharing network (Albery et al.~2020). This network stores pairwise
values of viral community similarity. To project viral sharing values
into a single value for every pixel, we averaged the pairwise scores.
High values of the average sharing propensity means that this specific
extant bat assemblage is likely to be proficient at exchanging viruses.

\hypertarget{composite-risk-map}{%
\subsection{Composite risk map}\label{composite-risk-map}}

To visualize the aggregated risk at the global scale, we combine the
three individual risk components (phylogenetic diversity, compositional
uniqueness, and viral sharing) using an additive color model (Seekell et
al.~2018). In this approach, every risk component gets assigned a
component in the RGB color model (phylogenetic diversity is green,
compositional uniqueness is red, and viral sharing is blue). In order to
achieve a valid RGB measure, all components are re-scaled to the
{[}0,1{]} interval. This additive model conveys both the intensity of
the overall risk, but also the nature of the risk as colors diverge
towards combinations of values for three risk components.

\hypertarget{viral-phylogeography}{%
\subsection{Viral phylogeography}\label{viral-phylogeography}}

We used the following query to pull all betacoronavirus sequence data
from the GenBank Nucleotide database except SARS-CoV-2;
(``Betacoronavirus''{[}Organism{]} OR betacoronavirus{[}All Fields{]})
NOT (``Severe acute respiratory syndrome coronavirus 2''{[}Organism{]}
OR sars-cov-2{[}All Fields{]}). We added a single representative
sequence for SARS-CoV-2 and manually curated to remove sequences without
the RNA-dependent RNA polymerase (RdRp) sequence or that contained words
indicating recombinant or laboratory strains including ``patent,''
``mutant,'' ``GFP,'' and ``recombinant.'' We filtered over-represented
taxa including betacoronavirus 1, hCoV-OC43, Middle East respiratory
syndrome coronavirus, Murine hepatitis virus, and hCoV-HKU1. Curated
betacoronavirus RdRp sequences were then aligned using MAFFT v 1.4.0
(Katoh and Standley 2013, Supplemental X) and a maximum likelihood tree
reconstructed in IQ-TREE v 1.6.12 (Nguyen et al.~2015) with ModelFinder
(Kalyaanamoorthy et al.~2017) ultrafast bootstrap approximation (Hoang
et al. 2018) and the following parameters (STEPH WILL ADD, Supplemental
X).

\hypertarget{viral-evolutionary-diversification}{%
\subsection{Viral evolutionary
diversification}\label{viral-evolutionary-diversification}}

We first tested the hypothesis that hotspots of viral diversification
would track hotspots of bat diversification. To do so, we plotted the
number of known bat hosts (specifically only those included in the
phylogeny, so there was a 1:1 correspondence between data sources)
against the ``mean evolutionary distinctiveness'' of the associated
viruses. To calculate this, we derived the fair proportions evolutionary
distinctiveness (Isaac et al., 2007) for each of the viruses in the
tree, then averaged these at the bat species level, projected these
values onto their geographic distributions, and averaged across every
bat found in a given pixel. As such, this can be thought of as a map of
the mean evolutionary distinctiveness of the known viral community
believed to be associated with a particular subset of bats present.

\hypertarget{co-distribution-of-hosts-and-viral-hotspots}{%
\subsection{Co-distribution of hosts and viral
hotspots}\label{co-distribution-of-hosts-and-viral-hotspots}}

Subsequently, we tested the hypothesis that the biogeography of bat
betacoronaviruses should track the biogeography of their hosts. To test
this idea, we loosely adapted a method from Kreft \& Jetz (2010), who
proposed a phylogenetic method for the delineation of animal
biogeographic regions. In their original method, a distance matrix -
where each row or column represents a geographic raster's grid cell, and
the dissimilarity values are the ``beta diversity similarity'' of their
community assemble - undergoes non-metric multidimensional scaling
(NMDS); the first two axes of the NMDS are projected geographically
using a four-color bivariate map.

Here, we build on this idea with an entirely novel methodology. First,
we measure the phylogenetic distance between the different viruses in
the betacoronavirus tree by using the cophenetic function in `ape';
subsequently, we take a principal components analysis of that distance
matrix (readily interchangeable for NMDS in this case) to project the
viral tree into an n-dimensional space. We then take the first two
principal components and, as with the evolutionary distinctiveness
analysis, aggregated these to a mean host value and projected them using
a four-color bivariate map.

\hypertarget{outbreaks-data-geo-referencing}{%
\subsection{Outbreaks data
geo-referencing}\label{outbreaks-data-geo-referencing}}

Finally, we provide a summary visualization of what available
information describes the spillover of zoonotic betacoronaviruses of bat
origin where data was available before and up through the COVID-19
pandemic. The SARS-CoV-2 outbreak was georeferenced to the initial case
cluster in Wuhan, China; SARS-CoV was georeferenced based on the cave
with the closest known viruses circulating in nature (Hu et al.~2017
PLoS Pathogens), and a nearby location where serological (antibody)
evidence has indicated human exposure to SARS-like viruses (Wang et
al.~2018 Virologica Sinica). For MERS-CoV, we presented the index cases
available from a recently-published compendium of MERS-CoV cases
(Ramshaw et al.~2019); these are largely if not all presumed to be
camel-to-human transmission, and the precise origin point of MERS-CoV in
bats is uncertain. Not shown is a recent case of a recombinant canine
coronavirus that showed the ability to infect humans, both because this
study was published after the beginning of the COVID-19 pandemic and
because bats' involvement in this cycle of transmission has been
marginal to non-existent.

\hypertarget{results}{%
\section{Results}\label{results}}
